---
title: "Downloading TCGA Diagnostic Images"
author:
  - name: Ilaria Billato
    affiliation: Department of Biology, University of Padova
date: "`r Sys.Date()`"
package: imageTCGAWorkflow
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Downloading TCGA Diagnostic Images}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

# Introduction

The TCGA diagnostic image database contains approximately **11,765 whole-slide
images (WSI)** in SVS format from ~9,640 patients across 32 cancer types.
Images are freely available from the
[NIH Genomic Data Commons (GDC)](https://portal.gdc.cancer.gov/).

Two complementary ways to find and download images are covered here:

1. **`imageTCGA` Shiny app** — interactive point-and-click interface with
   clinical filters and code generation
2. **`GenomicDataCommons` R package** — programmatic access for scripted or
   large-scale downloads

# Option 1: Shiny App (imageTCGA)

The `imageTCGA` app provides an interactive interface to the full TCGA image
catalogue. It is the recommended starting point for exploratory data selection.

```{r shiny, eval=FALSE}

imageTCGA::imageTCGA()
```

## App features

**Filter panel** — narrow down images by:

- Cancer type (project ID, e.g. TCGA-OV, TCGA-BRCA)
- Histological type
- Tumour grade and stage
- Patient sex and age range

**Dot plots** — available for gynaecological tumours (BRCA, OV, UCS, UCEC),
showing image counts stratified by clinical variables.

**Interactive maps** — display the geographic origin of each image at three
levels: institution (centre), US state, and country.

**Code generation panel** — once you are satisfied with the filter selection,
click *Generate code*. The app outputs a ready-to-run `GenomicDataCommons`
query (see Option 2 below) that you can copy directly into your R session.

# Option 2: GenomicDataCommons (programmatic)

The `GenomicDataCommons` Bioconductor package provides direct programmatic
access to the GDC REST API.

## Installation

```{r install-gdc}
BiocManager::install("GenomicDataCommons")
library(GenomicDataCommons)
```

## Query the image catalogue

```{r query}
# Authenticate (only needed for controlled-access data; images are open-access)
# gdc_token() # not required for diagnostic slides

# Build a query for diagnostic slides from ovarian cancer
q <- files() |>
    GenomicDataCommons::filter(
        ~ cases.project.project_id == "TCGA-OV" &
          data_type             == "Slide Image" &
          experimental_strategy == "Diagnostic Slide"
    )

# Preview results
res <- results(q, size = 10)
res[, c("file_id", "file_name", "file_size")]
```

## Filter by clinical metadata

```{r filter-clinical}
# Add clinical filters: stage IV, high-grade serous
q_filtered <- files() |>
    GenomicDataCommons::filter(
        ~ cases.project.project_id               == "TCGA-OV" &
          data_type                              == "Slide Image" &
          experimental_strategy                  == "Diagnostic Slide" &
          cases.diagnoses.tumor_stage            == "stage iv" &
          cases.diagnoses.morphology             == "8441/3"   # serous cystadenocarcinoma
    )

res_filtered <- results(q_filtered, size = 100)
nrow(res_filtered)
```

## Check available metadata fields

```{r fields}
# See all filterable fields for files
available_fields("files") |> grep("cases", x = _, value = TRUE) |> head(20)
```

## Download images

```{r download}
# Download a single image
file_ids <- results(q, size = 1)$file_id

gdcdata(
    uuids           = file_ids,
    destination_dir = "tcga_images/",
    progress        = TRUE
)
```

The downloaded SVS files can be very large (500 MB – 3 GB each). For
large-scale downloads, consider running on an HPC system and using the
[GDC Data Transfer Tool](https://gdc.cancer.gov/access-data/gdc-data-transfer-tool)
(command-line client).

## GDC Data Transfer Tool (batch download)

For hundreds of images, create a manifest file from the GDC portal or
via R and use the CLI tool:

```{r manifest}
# Write a manifest
manifest(q_filtered) |> write.table("gdc_manifest.txt",
    sep = "\t", quote = FALSE, row.names = FALSE)
```

```bash
# Then download with the GDC client
gdc-client download -m gdc_manifest.txt -d tcga_images/
```

## Metadata retrieval

Clinical and demographic metadata can be retrieved alongside file metadata:

```{r metadata}
# Get associated case metadata
case_ids <- res$cases |> sapply(\(x) x[[1]])

cases() |>
    GenomicDataCommons::filter(~ case_id %in% case_ids) |>
    select(c("case_id",
             "diagnoses.tumor_stage",
             "diagnoses.morphology",
             "demographic.gender",
             "demographic.age_at_index")) |>
    results(size = length(case_ids))
```

# Image Format

TCGA diagnostic images are stored as **SVS** (Aperio ScanScope) files, a
multi-resolution pyramidal TIFF format. In R, raw SVS files are typically
not opened directly — instead, pre-computed features (HoVerNet nuclei
segmentation, Prov-GigaPath embeddings) are accessed via `imageFeatureTCGA`
without requiring the original images.

If you need to inspect SVS files directly, use the
[openslide-python](https://openslide.org/api/python/) library:

```bash
# Install the system library and Python bindings
pip install openslide-bin openslide-python

# Inspect a slide in Python
python3 -c "
import openslide
slide = openslide.OpenSlide('tcga_images/TCGA-XX-XXXX.svs')
print('Levels:', slide.level_count)
print('Dimensions:', slide.dimensions)
print('MPP:', slide.properties.get('openslide.mpp-x'))
"
```

# Session Information

```{r session-info, eval=TRUE}
sessioninfo::session_info()
```
