---
title: "Reproducing HoVerNet Nuclei Segmentation on TCGA Images"
author:
  - name: Ilaria Billato
    affiliation: Department of Biology, University of Padova
date: "`r Sys.Date()`"
package: imageTCGAWorkflow
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Reproducing HoVerNet Nuclei Segmentation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

# Overview

This document describes how to reproduce the HoVerNet nuclei segmentation
results that underlie the pre-computed features available in
[imageFeatureTCGA](https://github.com/waldronlab/imageFeatureTCGA).

The complete setup instructions, scripts, and environment files are maintained
in the **[hovernethelp](https://github.com/hpages/hovernethelp)** repository
by Hervé Pagès.

HoVer-Net ([Graham et al., 2019](https://doi.org/10.1016/j.media.2019.101563))
performs simultaneous nuclei **detection**, **segmentation**, and **classification**
on histopathology whole-slide images (WSI). The pipeline was run on all TCGA
diagnostic images and the outputs (JSON, GeoJSON, H5AD, thumbnails) are
accessible via `imageFeatureTCGA`.

# Hardware Requirements

Running HoVerNet from scratch requires GPU resources:

| Resource | Minimum |
|---|---|
| GPU | NVIDIA A100 or L40S |
| CPUs | 16 |
| RAM | 60 GB |
| OS | Ubuntu 24.04 |
| Storage | ~780 GB for full TCGA outputs |

Processing time is approximately **76–81 minutes per image** on a single GPU node.
A 7-node mini-cluster (described below) achieves the full TCGA dataset in
15–25 hours.

> If you only need the pre-computed results, skip to the
> [imageFeatureTCGA workflow vignette](imageTCGAWorkflow.html) — no GPU is
> required to access the stored features.

# Environment Setup

All instructions below follow the
[hovernethelp setup guide](https://github.com/hpages/hovernethelp/blob/main/setup_hovernet_Ubuntu2404.txt).

## GPU drivers

For **NVIDIA A100**:
```bash
sudo apt-get install nvidia-linux-grid-535
```

For **NVIDIA L40S**:
```bash
sudo apt-get install nvidia-driver-550
```

Verify the installation:
```bash
nvidia-smi
```

## Miniconda and HoVer-Net

```bash
# Install Miniconda
wget https://repo.anaconda.com/miniconda/Miniconda3-latest-Linux-x86_64.sh
bash Miniconda3-latest-Linux-x86_64.sh
conda init bash && source ~/.bashrc

# Clone HoVer-Net
git clone https://github.com/vqdang/hover_net
cd hover_net

# Create conda environment from hovernethelp
wget https://raw.githubusercontent.com/hpages/hovernethelp/main/environment.yml
conda env create -f environment.yml
conda activate hovernet
```

The `environment.yml` pins key packages:

- Python 3.12.3
- PyTorch 2.5.1
- `opencv-python`, `scikit-image`, `scikit-learn`, `scipy`, `pandas`
- `openslide` for whole-slide image reading

## Pre-trained model

The pannuke_type model (TF2→PyTorch conversion) is downloaded with `gdown`:

```bash
pip install gdown
gdown <model_gdrive_id> -O pannuke_type_tf2pytorch.tar
tar -xf pannuke_type_tf2pytorch.tar
```

See the [hovernethelp setup file](https://github.com/hpages/hovernethelp/blob/main/setup_hovernet_Ubuntu2404.txt)
for the exact model ID and path configuration.

# Downloading TCGA Images

Images are downloaded using the `GenomicDataCommons` Bioconductor package.
The `imageTCGA` Shiny app generates the query code interactively (see the
[main workflow vignette](imageTCGAWorkflow.html#step-1-data-discovery-with-imagetcga)).

```r
# Example: download a single OV diagnostic image
library(GenomicDataCommons)

q <- files() |>
    filter(~ cases.project.project_id == "TCGA-OV" &
               data_type == "Slide Image" &
               experimental_strategy == "Diagnostic Slide") |>
    results(size = 1)

gdcdata(q$file_id, destination_dir = "tcga_images/")
```

# Running HoVerNet Inference

## Single image

```bash
python run_infer.py \
  --gpu='0' \
  --nr_types=6 \
  --type_info_path=type_info.json \
  --batch_size=12 \
  --model_mode=original \
  --model_path=pannuke_type_tf2pytorch/hovernet_fast_pannuke_type_tf2pytorch.tar \
  --nr_inference_workers=1 \
  --nr_post_proc_workers=3 \
  wsi \
  --input_dir=tcga_images/ \
  --output_dir=hovernet_output/ \
  --save_thumb \
  --save_mask
```

Key parameters:

| Parameter | Value | Note |
|---|---|---|
| `nr_types` | 6 | Nuclei classes (neoplastic, inflammatory, connective, dead, epithelial, background) |
| `batch_size` | 12 | Adjust to GPU memory |
| `nr_inference_workers` | 1 | GPU workers |
| `nr_post_proc_workers` | 3 | CPU post-processing workers |

## Batch processing (multiple images)

The
[`infer_batch2.sh`](https://github.com/hpages/hovernethelp/blob/main/infer_batch2.sh)
script automates the full pipeline:

1. Reads a manifest of TCGA image IDs to process
2. Downloads each SVS file via `GenomicDataCommons` (with rsync retry, up to
   10 attempts with 5-minute intervals)
3. Runs HoVerNet inference and validates JSON output
4. Pushes results to a central storage node (`hoverboss`)
5. Updates the manifest with success/failure status and timestamps

```bash
# Usage
bash infer_batch2.sh manifest.tsv
```

# Output Files

For each input SVS image, HoVerNet produces:

| File | Format | Content |
|---|---|---|
| `<id>.json` | JSON | Per-nucleus centroid, contour, type, probabilities |
| `<id>.geojson` | GeoJSON | Same data in GIS-compatible format |
| `<id>.h5ad` | H5AD | Feature matrix (intensity, morphology statistics) |
| `<id>.png` | PNG | Tissue thumbnail with segmentation overlay |

These files are exactly what `imageFeatureTCGA` imports via `HoverNet() |> import()`
— see the [imageFeatureTCGA package page](pkg-imageFeatureTCGA.html) for
how to load them into a `SpatialExperiment`.

# Scaling with a Mini-Cluster

The [hovernethelp minicluster setup](https://github.com/hpages/hovernethelp/tree/main/minicluster)
describes a 7-node architecture:

- **Worker nodes** (`hovernet1–4`, `hovernet6–7`, `kakapo1`): GPU nodes running inference
- **Central node** (`hoverboss`): 4 CPUs, 779 GB storage, collects results

Workers sync completed JSON/H5AD files to `hoverboss` via cron every 15 minutes.
Each worker processes a non-overlapping range of the TCGA image list. Total
throughput: **100+ images/day**, completing the full TCGA dataset in 15–25 hours.

# Session Information

```{r session-info, eval=TRUE}
sessioninfo::session_info()
```
