---
title: "Histopathology Image Analysis with TCGA Data: an end-to-end workflow"
author:
  - name: Ilaria Billato
    affiliation: Department of Biology, University of Padova
  - name: Levi Waldron
    affiliation: CUNY Graduate School of Public Health and Health Policy
date: "`r Sys.Date()`"
package: imageTCGAWorkflow
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Histopathology Image Analysis Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  cache     = TRUE,
  message   = FALSE,
  warning   = FALSE
)
```

# Introduction

This workflow demonstrates a complete analysis pipeline for histopathology
image data from The Cancer Genome Atlas (TCGA), using a suite of Bioconductor
packages developed around the `imageTCGA` ecosystem:

| Package | Role |
|---|---|
| [imageTCGA](https://github.com/billila/imageTCGA) | Interactive Shiny app to explore and select TCGA diagnostic images |
| [imageFeatureTCGA](https://github.com/waldronlab/imageFeatureTCGA) | Import HoVerNet and Prov-GigaPath features into Bioconductor objects |
| [imageTCGAutils](https://github.com/waldronlab/imageTCGAutils) | Spatial statistics and dimensionality reduction on tile/cell data |
| [HistoImagePlot](https://github.com/waldronlab/HistoImagePlot) | Overlay cell segmentation on tissue thumbnails |

The TCGA image database contains ~11,765 diagnostic whole-slide images (WSI)
from ~9,640 patients across many cancer types. Pre-computed features are
available for 54,253 files: 33,177 from HoVerNet (nuclei segmentation/
classification) and 21,076 from Prov-GigaPath (slide- and tile-level deep
learning embeddings).

## Installation

These packages are available from GitHub during active development and will be
submitted to Bioconductor. Install them as follows:

```{r install, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

# Bioconductor dependencies
BiocManager::install(c(
    "SpatialExperiment", "SingleCellExperiment", "MultiAssayExperiment",
    "BiocFileCache", "BiocIO", "BumpyMatrix", "GenomicDataCommons"
))

# imageTCGA ecosystem (GitHub)
BiocManager::install(c(
    "billila/imageTCGA",
    "waldronlab/imageFeatureTCGA",
    "waldronlab/imageTCGAutils",
    "waldronlab/HistoImagePlot"
))
```

## Load packages

```{r load-pkgs, eval=FALSE}
library(imageTCGA)
library(imageFeatureTCGA)
library(imageTCGAutils)
library(HistoImagePlot)
library(MultiAssayExperiment)
library(SpatialExperiment)
library(dplyr)
```

---

# Step 1: Data Discovery with imageTCGA

The `imageTCGA` Shiny application provides an interactive interface to the
TCGA diagnostic image database. It allows filtering by cancer type, clinical
parameters (histological type, grade, stage), and patient geography, and
generates ready-to-run `GenomicDataCommons` download code.

```{r shiny-app, eval=FALSE}
# Launch the Shiny app
imageTCGA::run_app()
```

The app displays:

- **Dot plots** of image counts stratified by clinical variables for
  gynaecological tumours (BRCA, OV, UCS, UCEC).
- **Interactive maps** showing the origin institution of each image at
  country, US-state, and centre level.
- A **code generation panel** that produces the `GenomicDataCommons` query
  corresponding to the current filter selection.

Once you have identified the case(s) of interest, copy the generated code for
use in the next step.

---

# Step 2: Retrieving Pre-computed Features with imageFeatureTCGA

`imageFeatureTCGA` provides access to a catalogue of 54,253 pre-computed
feature files hosted on the Waldron Lab server. Two computational pipelines
are covered:

- **HoVerNet** – nuclei detection, segmentation, and classification at
  single-nucleus resolution. Output formats: JSON (coordinates + contours),
  GeoJSON, H5AD (features), and PNG thumbnails.
- **Prov-GigaPath** – vision foundation model embeddings at slide level
  (one vector per WSI) and tile level (one vector per 256 × 256 px tile).

## Browse the data catalogue

```{r catalogue}
# Inspect the full data catalogue
cat <- imageFeatureTCGA::imageCatalogue()
head(cat)
```

The catalogue columns include `file_id`, `cases.submitter_id` (TCGA barcode),
`cancer_type`, `pipeline`, `data_type`, and the remote `url`.

```{r catalogue-filter}
# Filter to HoVerNet H5AD files for ovarian cancer
ov_h5ad <- cat |>
    dplyr::filter(cancer_type == "OV",
                  pipeline == "HoVerNet",
                  data_type == "h5ad")

nrow(ov_h5ad)
head(ov_h5ad[, c("cases.submitter_id", "pipeline", "data_type")])
```

## Import into a SpatialExperiment (HoVerNet)

HoVerNet outputs are imported as `SpatialExperiment` objects where each row
is a nucleus and spatial coordinates reflect its centroid on the slide.

```{r spe-hovernet, eval=FALSE}
# Import HoVerNet H5AD data for a single slide
spe_hover <- imageFeatureTCGA::importHoVerNet(
    file_id  = ov_h5ad$file_id[1],
    format   = "h5ad"
)
spe_hover
```

```{r spe-hovernet-json, eval=FALSE}
# Import HoVerNet JSON (includes contours) for the same slide
spe_json <- imageFeatureTCGA::importHoVerNet(
    file_id  = ov_h5ad$file_id[1],
    format   = "json"
)
```

The resulting `SpatialExperiment` contains:
- `assay("X")` – per-nucleus feature matrix (intensity statistics, morphology)
- `spatialCoords()` – centroid (x, y) in slide pixel coordinates
- `colData()` – cell type label, nucleus area, and QC flags

## Import tile embeddings (Prov-GigaPath)

```{r spe-gigapath, eval=FALSE}
# Filter catalogue to tile-level Prov-GigaPath files for one OV slide
pg_tiles <- cat |>
    dplyr::filter(cases.submitter_id == ov_h5ad$cases.submitter_id[1],
                  pipeline == "ProvGigaPath",
                  data_type == "tile_embeddings")

spe_tiles <- imageFeatureTCGA::importProvGigaPath(
    file_id  = pg_tiles$file_id[1],
    level    = "tile"
)
spe_tiles
```

Each column is a 256 × 256 px tile; the 1,536-dimensional embedding vector
from Prov-GigaPath is stored in `assay("embeddings")`.

## Assemble a MultiAssayExperiment

Both assays can be linked to TCGA clinical metadata in a single
`MultiAssayExperiment`:

```{r mae, eval=FALSE}
mae <- MultiAssayExperiment::MultiAssayExperiment(
    experiments = list(
        HoVerNet    = spe_hover,
        GigaPath    = spe_tiles
    )
)
mae
```

---

# Step 3: Spatial Analysis with imageTCGAutils

`imageTCGAutils` provides utility functions for dimensionality reduction and
spatial autocorrelation on the tile- and nucleus-level data.

## PCA on tile embeddings

```{r pca, eval=FALSE}
# Run PCA on Prov-GigaPath tile embeddings and store result in reducedDim
spe_tiles <- imageTCGAutils::runTilePCA(spe_tiles, ncomponents = 30)

# Visualise first two PCs coloured by tissue region
imageTCGAutils::plotTileReducedDim(spe_tiles, dimred = "PCA",
                                    colour_by = "tissue_class")
```

## Spatial autocorrelation

Moran's I and Geary's C quantify whether a feature (e.g. proportion of
tumour nuclei in a tile) is spatially clustered, dispersed, or random.
Local Moran's I (LISA) identifies individual tiles that drive global clustering.

```{r morans, eval=FALSE}
# Compute global Moran's I for the tumour-nucleus proportion per tile
mi <- imageTCGAutils::globalMoransI(
    spe   = spe_tiles,
    assay = "embeddings",
    feat  = "tumour_prop"
)
mi

# Local Moran's I (LISA) — returns a SpatialExperiment with LISA scores
spe_tiles <- imageTCGAutils::localMoransI(spe_tiles, feat = "tumour_prop")

# Geary's C
gc <- imageTCGAutils::globalGearysC(spe_tiles, feat = "tumour_prop")
gc
```

## Matching HoVerNet nuclei to Prov-GigaPath tiles

```{r match, eval=FALSE}
# Assign each nucleus (spe_hover) to its containing tile (spe_tiles)
spe_hover <- imageTCGAutils::matchNucleiToTiles(
    nuclei = spe_hover,
    tiles  = spe_tiles
)
# colData(spe_hover)$tile_id now contains the parent tile identifier
```

---

# Step 4: Visualization with HistoImagePlot

`HistoImagePlot` renders side-by-side panels of the tissue thumbnail and
nucleus segmentation overlay, useful for quality control and for illustrating
findings.

## Side-by-side thumbnail + segmentation

```{r histoplot, eval=FALSE}
# Import HoVerNet JSON for the slide (includes nucleus contours)
spe_json <- HistoImagePlot::importHoVerNetJSON(
    file_id = ov_h5ad$file_id[1]
)

# Plot thumbnail next to coloured segmentation map
HistoImagePlot::plotHistoImage(
    spe        = spe_json,
    colour_by  = "cell_type",
    point_size = 0.4,
    title      = paste("OV –", ov_h5ad$cases.submitter_id[1])
)
```

Supported colour palettes can be passed via `palette =`. The thumbnail is
fetched from the catalogue and cached automatically by `BiocFileCache`.

## Overlay on a region of interest

```{r roi, eval=FALSE}
# Restrict to a 2000 × 2000 px region starting at (5000, 3000)
HistoImagePlot::plotHistoImage(
    spe       = spe_json,
    colour_by = "cell_type",
    xlim      = c(5000, 7000),
    ylim      = c(3000, 5000)
)
```

---

# Session Information

```{r session-info}
sessioninfo::session_info()
```
