---
title: "Package: HistoImagePlot"
author:
  - name: Ilaria Billato
    affiliation: Department of Biology, University of Padova
date: "`r Sys.Date()`"
package: imageTCGAWorkflow
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Package: HistoImagePlot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

# Overview

**GitHub:** <https://github.com/waldronlab/HistoImagePlot>

`HistoImagePlot` provides visualization tools for histopathology research,
working alongside `imageFeatureTCGA` to import and display HoVerNet nuclei
segmentation results overlaid on tissue thumbnail images.

Supported HoVerNet input formats:

| Format | Extension | Content |
|---|---|---|
| H5AD | `.h5ad` / `.h5ad.gz` | Feature matrix + spatial statistics (mean intensity, nearest-neighbour distance) |
| JSON | `.json` / `.json.gz` | Nucleus centroids, contours, type, probabilities |

```{r load}
library(imageFeatureTCGA)
library(HistoImagePlot)
library(dplyr)
library(SpatialExperiment)
library(ggplot2)
library(cowplot)
```

# Import HoVerNet H5AD Data

Import a HoVerNet H5AD file into a `SpatialExperiment` object.
The file is automatically cached via `BiocFileCache`:

```{r import}
hov_file <- paste0(
    "https://store.cancerdatasci.org/hovernet/h5ad/",
    "TCGA-23-1021-01Z-00-DX1.F07C221B-D401-47A5-9519-10DE59CA1E9D.h5ad.gz"
)

hn_spe <- HoverNet(hov_file, outClass = "SpatialExperiment") |>
    import()

hn_spe
```

The corresponding tissue thumbnail is also available from the same store:

```{r thumb}
thumb_path <- paste0(
    "https://store.cancerdatasci.org/hovernet/thumb/",
    "TCGA-23-1021-01Z-00-DX1.F07C221B-D401-47A5-9519-10DE59CA1E9D.png"
)
```

# Overlaying Segmentation on Tissue Thumbnails

## Basic Overlay

`plotHoverNetH5ADOverlay()` renders nucleus centroids coloured by cell type
on top of the tissue thumbnail:

```{r overlay-basic}
plotHoverNetH5ADOverlay(hn_spe, thumb_path)
```

## Customised Overlay

Point size and legend appearance can be adjusted:

```{r overlay-custom}
plotHoverNetH5ADOverlay(
    hn_spe,
    thumb_path,
    title             = "Ovarian Cancer Tissue - Cell Segmentation",
    point_size        = 0.02,
    legend_point_size = 3
)
```

## Custom Colour Palette

Pass a named character vector to override the default cell-type colours:

```{r overlay-colors}
custom_colors <- c(
    "no label"          = "#808080",
    "neoplastic"        = "#E31A1C",
    "inflammatory"      = "#1F78B4",
    "stromal"           = "#33A02C",
    "necrotic"          = "#FF7F00",
    "benign epithelial" = "#6A3D9A"
)

plotHoverNetH5ADOverlay(
    hn_spe,
    thumb_path,
    color_palette = custom_colors,
    title         = "Custom Color Scheme"
)
```

# Visualising H5AD Features

H5AD files include additional per-nucleus statistics stored in `colData()`:
`mean_intensity` and `nearest_neighbor_distance`. Spatial coordinates are
available via `spatialCoords()` as `x_centroid` and `y_centroid`.

```{r h5ad-features}
h5ad_coords <- data.frame(spatialCoords(hn_spe), colData(hn_spe))

p1 <- ggplot(h5ad_coords, aes(x = x_centroid, y = y_centroid,
                               color = mean_intensity)) +
    geom_point(size = 0.5) +
    scale_color_viridis_c() +
    coord_fixed() +
    theme_minimal() +
    labs(title = "Mean Intensity", color = "Intensity")

p2 <- ggplot(h5ad_coords, aes(x = x_centroid, y = y_centroid,
                               color = nearest_neighbor_distance)) +
    geom_point(size = 0.5) +
    scale_color_viridis_c(option = "plasma") +
    coord_fixed() +
    theme_minimal() +
    labs(title = "Nearest Neighbor Distance", color = "Distance")

cowplot::plot_grid(p1, p2, ncol = 2)
```

# Session Information

```{r session-info, eval=TRUE}
sessioninfo::session_info()
```
