---
title: "Package: imageTCGAutils"
author:
  - name: Ilaria Billato
    affiliation: Department of Biology, University of Padova
date: "`r Sys.Date()`"
package: imageTCGAWorkflow
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Package: imageTCGAutils}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

# Overview

**GitHub:** <https://github.com/waldronlab/imageTCGAutils>

`imageTCGAutils` provides utility functions for integrating and analyzing
multi-modal whole-slide image (WSI) data from TCGA. It is designed to work
alongside `imageFeatureTCGA`, which handles import of pre-computed features
from HoVerNet and Prov-GigaPath.

```{r load}
library(imageFeatureTCGA)
library(imageTCGAutils)
library(ggplot2)
library(dplyr)
library(sfdep)
library(spdep)
library(SpatialExperiment)
library(data.table)
```

# Import Prov-GigaPath Tile-Level Embeddings

```{r import-tiles}
# Select an ovarian cancer slide as example
tile_prov_url <- paste0(
    "https://store.cancerdatasci.org/provgigapath/tile_level/",
    "TCGA-23-1021-01Z-00-DX1.F07C221B-D401-47A5-9519-10DE59CA1E9D.csv.gz"
)

example_slide <- ProvGiga(tile_prov_url) |>
    import()
```

# PCA on Tile Embeddings

```{r pca}
# Extract embedding columns (named with numbers: "0", "1", ...)
embedding_cols <- grep("^[0-9]+$", names(example_slide), value = TRUE)

# Run PCA
pca_res <- prcomp(example_slide[, embedding_cols], scale. = TRUE)

pca_example_slide <- bind_cols(
    example_slide,
    as_tibble(pca_res$x)[, 1:2]
)

# PCA scatter plot
ggplot(pca_example_slide, aes(PC1, PC2)) +
    geom_point(alpha = 0.6, size = 1) +
    theme_minimal() +
    labs(title = "Tile-level PCA Ovarian Cancer Embedding: Single Slide")

# Spatial layout coloured by PC1
ggplot(pca_example_slide, aes(tile_x, tile_y, color = PC1)) +
    geom_point(size = 1) +
    scale_color_viridis_c() +
    coord_equal() +
    theme_minimal() +
    labs(title = "Tissue layout colored by PC1")
```

# Spatial Autocorrelation

To investigate spatial patterns, we build a k-nearest neighbour graph
from tile coordinates and compute global and local spatial autocorrelation
on the PCA-reduced embeddings.

```{r spatial-weights}
coords <- pca_example_slide[, c("tile_x", "tile_y")]
nb <- knn2nb(knearneigh(coords, k = 6))
lw <- nb2listw(nb, style = "W")
```

## Global Moran's I and Geary's C

```{r global}
mi <- moran.test(pca_example_slide$PC1, lw)
gc <- geary.test(pca_example_slide$PC1, lw)

mi
gc
```

Moran's I ∈ [−1, 1]: positive values indicate spatial clustering.
Geary's C ∈ [0, 2]: values < 1 indicate positive spatial autocorrelation.

## Local Moran's I (LISA)

```{r lisa}
lisa <- localmoran(pca_example_slide$PC1, lw)
pca_example_slide$localI      <- lisa[, "Ii"]
pca_example_slide$localI_pval <- lisa[, "Pr(z != E(Ii))"]

# Moran scatterplot
moran.plot(pca_example_slide$PC1, lw, labels = FALSE,
           main = "Moran scatterplot of PC1")

# LISA spatial map
ggplot(pca_example_slide, aes(x = tile_x, y = tile_y, color = localI)) +
    geom_point(size = 0.5) +
    scale_color_viridis_c() +
    coord_equal() +
    theme_minimal() +
    ggtitle("Local Moran's I (LISA) for PC1")
```

# Integration with HoVerNet Nuclei

`matchHoverNetToTiles()` computes the scaling factor between HoVerNet nucleus
coordinates and Prov-GigaPath tile coordinates, then assigns each tile a
dominant cell type and nucleus count based on the overlapping nuclei.

```{r import-hover}
hov_file <- paste0(
    "https://store.cancerdatasci.org/hovernet/h5ad/",
    "TCGA-23-1021-01Z-00-DX1.F07C221B-D401-47A5-9519-10DE59CA1E9D.h5ad.gz"
)

hn_spe <- HoverNet(hov_file, outClass = "SpatialExperiment") |>
    import()
```

```{r plot-coords}
# Visualise nucleus vs tile coordinates — they use different scales
cell_coords <- spatialCoords(hn_spe)

plot(cell_coords[, 1], cell_coords[, 2],
     pch = 16, col = "#0000FF20",
     main = "HoVerNet nuclei (blue) vs GigaPath tiles (red)")
points(pca_example_slide$tile_x, pca_example_slide$tile_y,
       pch = 16, col = "#FF000020")
```

```{r match}
pg_spe <- ProvGiga(tile_prov_url) |> import()

match_hv_pg <- matchHoverNetToTiles(hn_spe, pg_spe)

# Dominant cell type per tile
ggplot(match_hv_pg$tiles_with_nuclei,
       aes(tile_x, tile_y, color = dominant_cell_type)) +
    geom_point(size = 2) +
    coord_equal() +
    theme_minimal() +
    labs(title = "Per-tile dominant HoVerNet cell type")

# All cell types per tile (sized by count)
ggplot(match_hv_pg$tiles_with_nuclei,
       aes(tile_x, tile_y, color = cell_type_label, size = N)) +
    geom_point(alpha = 0.7) +
    coord_equal() +
    theme_minimal() +
    labs(title = "All HoVerNet cell types per tile")
```

# Session Information

```{r session-info, eval=TRUE}
sessioninfo::session_info()
```
