---
title: "Package: imageTCGA"
author:
  - name: Ilaria Billato
    affiliation: Department of Biology, University of Padova
date: "`r Sys.Date()`"
package: imageTCGAWorkflow
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Package: imageTCGA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

# Overview

**GitHub:** <https://github.com/billila/imageTCGA>

`imageTCGA` is a Shiny application for interactive exploration of the
TCGA diagnostic image database. It provides clinical and geographical
filtering of ~11,765 images from ~9,640 patients, and generates
ready-to-run `GenomicDataCommons` R code for downloading selected images.

# Launch the app

```{r run}
library(imageTCGA)
imageTCGA::imageTCGA()
```

# Features

## Clinical filtering

The left panel allows filtering by:

- **Project** (cancer type, e.g. TCGA-OV, TCGA-BRCA)
- **Histological type** (morphology code)
- **Tumour grade** and **stage**


## Dot plots

For gynaecological tumours (BRCA, OV, UCS, UCEC), dot plots show the
distribution of images across clinical strata.

## Interactive maps

Images are annotated with the contributing institution. The map tab
displays image origins at three levels:

- Institution (centre name)
- US state
- Country

## Code generation

After applying filters, the **Generate code** panel produces a
`GenomicDataCommons` query that reproduces the current selection:

```{r generated-code-example}
# Example of generated code:
library(GenomicDataCommons)
files() |>
    filter(
        ~ cases.project.project_id == "TCGA-OV" &
          data_type == "Slide Image" &
          experimental_strategy == "Diagnostic Slide"
    ) |>
    results(size = 100)
```

Copy this code into your R session, then pass the `file_id` column to
`gdcdata()` to download the SVS files.

# Session Information

```{r session-info, eval=TRUE}
sessioninfo::session_info()
```
