---
title: "Package: imageFeatureTCGA"
author:
  - name: Ilaria Billato
    affiliation: Department of Biology, University of Padova
date: "`r Sys.Date()`"
package: imageTCGAWorkflow
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Package: imageFeatureTCGA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

# Overview

**GitHub:** <https://github.com/waldronlab/imageFeatureTCGA>

`imageFeatureTCGA` provides convenient access to histopathology-derived
features from TCGA through two complementary pipelines:

- **HoVerNet** — cell segmentation and classification at single-nucleus level
- **ProvGigaPath** — slide- and tile-level deep learning embeddings

The data catalogue contains **54,253 pre-computed files**: 33,177 HoVerNet
files and 21,076 ProvGigaPath files. All files can be imported directly into
Bioconductor objects without requiring local image files or GPU resources.

```{r load}
library(imageFeatureTCGA)
library(SummarizedExperiment)
library(dplyr)
```

# Data Catalogue

Use `getCatalog()` to download the full catalogue of available files:

```{r catalogue}
getCatalog()
```

Filter by pipeline:

```{r catalogue-hover}
# HoVerNet files (33,177)
getCatalog("hovernet")

# ProvGigaPath files (21,076)
getCatalog("provgigapath")
```

Key columns in the catalogue: `pipeline`, `format`, `filename`, `tcga_barcode`,
`Case.ID`, `Project.ID`, `level`, `lat`, `lon`.

**HoVerNet formats:** `json`, `h5ad`, `thumb` (PNG thumbnail)

**ProvGigaPath formats:** `csv` (slide-level and tile-level, distinguished by
the `level` column: `"slide_level"` or `"tile_level"`)

# Importing HoVerNet Data

HoVerNet segmentation results are imported as `SpatialExperiment` or
`SpatialFeatureExperiment`. Columns represent individual nuclei.

```{r hover-import}
hspe <- getCatalog("hovernet") |>
    dplyr::filter(
        filename == paste(
            "TCGA-VG-A8LO-01A-01-DX1",
            "B39A4D64-82A1-4A04-8AB6-918F3058B83B",
            "json", "gz", sep = "."
        )
    ) |>
    getFileURLs() |>
    HoverNet(outClass = "SpatialExperiment") |>
    import()

hspe
#> class: SpatialExperiment
#> dim: 0 67081
#> metadata(1): type_map
#> assays(1): counts
#> colData names(10): cell_id x ... B sample_id
#> spatialCoords names(2): x y
```

Each nucleus has:

- `x`, `y` spatial coordinates (pixels relative to the slide)
- `type` — integer cell type code
- `type_prob` — classification probability
- `label` — cell type name (e.g. `"neopla"`, `"inflam"`, `"connec"`, `"necros"`)

```{r hover-coldata}
colData(hspe)
```

# Importing ProvGigaPath Embeddings

## Slide-level embeddings

Each row is a slide; the embedding vector (768 dimensions, named `V1`…`V768`)
summarises the entire WSI. `ProvGiga()` for a single file returns a tibble.

```{r slide-level}
getCatalog("provgigapath") |>
    dplyr::filter(
        filename == paste(
            "TCGA-VG-A8LO-01A-01-DX1",
            "B39A4D64-82A1-4A04-8AB6-918F3058B83B",
            "csv", "gz", sep = "."
        ),
        level == "slide_level"
    ) |>
    getFileURLs() |>
    ProvGiga() |>
    import()
#> # A tibble: 1 × 771
#>   slideName  tumorType fileName     V1     V2     V3  ...
#>   <chr>      <chr>     <chr>     <dbl>  <dbl>  <dbl>
#> 1 TCGA-VG-…  <NA>      TCGA-VG… -0.355 0.584 -0.402 …
#> # ℹ 768 embedding columns (V1–V768) + 3 metadata columns
```

## Tile-level embeddings

Each row is a 256 × 256 px tile; columns include `tile_x`, `tile_y`, and
1,536 embedding dimensions (`0`…`1535`).

```{r tile-level}
getCatalog("provgigapath") |>
    dplyr::filter(
        filename == paste(
            "TCGA-VG-A8LO-01A-01-DX1",
            "B39A4D64-82A1-4A04-8AB6-918F3058B83B",
            "csv", "gz", sep = "."
        ),
        level == "tile_level"
    ) |>
    getFileURLs() |>
    ProvGiga() |>
    import()
#> # A tibble: 211 × 1,543
#>   slide_name  tile_id tile_name tile_x tile_y    `0`    `1`  ...
#>   <chr>         <dbl> <chr>      <dbl>  <dbl>  <dbl>  <dbl>
#> 1 TCGA-VG-…        0 02155x_…   2155  18612  0.177  -1.42  …
#> # ℹ 201 more rows; 1,536 embedding columns (named `0`–`1535`) + tile_x, tile_y
```

# Importing Multiple Files

`ProvGigaList()` accepts multiple files at once. When all files are the same
level, `import()` returns a single `SummarizedExperiment`; for mixed levels it
returns a named list.

```{r multi-slide}
# Three slide-level files for TCGA-GBM → single SummarizedExperiment (768 × 3)
pgl <- getCatalog("provgigapath") |>
    dplyr::filter(level == "slide_level", Project.ID == "TCGA-GBM") |>
    dplyr::slice(1:3) |>
    getFileURLs() |>
    ProvGigaList() |>
    import()

pgl
#> class: SummarizedExperiment
#> dim: 768 3
#> metadata(5): slideName tumorType fileName patientIds sampleIds
#> assays(1): embeddings
#> colnames(3): TCGA-02-0001-01Z-00-DX1 TCGA-02-0001-01Z-00-DX2 TCGA-02-0001-01Z-00-DX3
```

```{r mixed-levels}
# Both levels for the same slide → list with $slide_level and $tile_level
pgl_mixed <- getCatalog("provgigapath") |>
    dplyr::filter(
        filename == paste(
            "TCGA-VG-A8LO-01A-01-DX1",
            "B39A4D64-82A1-4A04-8AB6-918F3058B83B",
            "csv", "gz", sep = "."
        ),
        level %in% c("slide_level", "tile_level")
    ) |>
    getFileURLs() |>
    ProvGigaList() |>
    import()

pgl_mixed$slide_level
#> class: SummarizedExperiment
#> dim: 768 1
#> assays(1): embeddings
#> colnames(1): TCGA-VG-A8LO-01A-01-DX1

pgl_mixed$tile_level
#> class: SummarizedExperiment
#> dim: 1538 1
#> assays(1): tiles
#> colnames(1): TCGA-VG-A8LO-01A-01-DX1
```


# Session Information

```{r session-info, eval=TRUE}
sessioninfo::session_info()
```
