---
title: "Prov-GigaPath Feature Extraction"
author:
  - name: Ilaria Billato
    affiliation: Department of Biology, University of Padova
date: "`r Sys.Date()`"
package: imageTCGAWorkflow
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Prov-GigaPath Feature Extraction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

# Overview

**Prov-GigaPath** ([Xu et al., 2024](https://doi.org/10.1038/s41586-024-07441-w))
is a whole-slide pathology foundation model trained on over 1.3 billion
pathology image tiles from Providence Health & Services. It produces
high-quality, general-purpose embeddings for computational pathology.

We ran Prov-GigaPath on all available TCGA diagnostic WSIs and extracted
embeddings at two granularities:

| Level | Unit | Embedding dim | Files in catalogue |
|---|---|---|---|
| **Tile level** | 256 × 256 px tile | 1,536 | ~14,000 CSV files |
| **Slide level** | Entire WSI (aggregated) | 1,536 | ~7,000 CSV files |

The resulting features are accessible without any GPU via
`imageFeatureTCGA` — see the
[main workflow vignette](imageTCGAWorkflow.html) for import examples.

This document describes how the extraction was performed, for users who
wish to re-run or extend the pipeline.

# Model

Prov-GigaPath uses a two-stage architecture:

1. **Tile encoder** — a DINOv2 ViT-g/14 model fine-tuned on 1.3 B pathology
   tiles. Each 256 × 256 px tile at 0.5 µm/px is encoded into a 1,536-dimensional
   vector.
2. **Slide encoder** — a LongNet transformer that aggregates tile-level tokens
   into a single slide-level representation, preserving spatial context via
   dilated attention.

The model weights are distributed via Hugging Face
(`prov-gigapath/prov-gigapath`) under the
[CC-BY-NC-4.0 license](https://creativecommons.org/licenses/by-nc/4.0/).
Access requires a Hugging Face account and accepting the model terms.

# Container

Inference was run inside an Apptainer/Singularity container to ensure
reproducibility across different compute environments.

## Definition file

```singularity
Bootstrap: docker
From: mambaorg/micromamba:1.5.8

%files
    environment.yaml /environment.yaml
    . /app

%post
    export MAMBA_ROOT_PREFIX=/opt/conda
    export PATH=$MAMBA_ROOT_PREFIX/bin:$PATH

    apt-get update && apt-get install -y git && apt-get clean

    # Create the gigapath conda environment
    micromamba create -y -n gigapath -f /environment.yaml

    # Install the package and dependencies
    cd /app
    micromamba run -n gigapath pip install -e .
    micromamba run -n gigapath pip install "timm>=1.0.3"
    micromamba run -n gigapath pip install "openslide-bin"

    micromamba clean --all --yes

%environment
    export MAMBA_ROOT_PREFIX=/opt/conda
    export PATH=$MAMBA_ROOT_PREFIX/envs/gigapath/bin:$PATH
```

> **Note:** The Hugging Face token (`HF_TOKEN`) is required to download the
> model weights and must be set as an environment variable at runtime — do
> **not** embed it in the container definition or commit it to version control.
> Set it at runtime: `apptainer run --env HF_TOKEN=<your_token> gigapath.sif`

## Build the container

```bash
apptainer build gigapath.sif gigapath.def
```

# Running Tile-Level Extraction

Tiles are extracted at **0.5 µm/px** (20× equivalent magnification),
tiled into non-overlapping 256 × 256 px patches, and encoded by the
DINOv2 tile encoder.

```python
import gigapath
import torch
from PIL import Image
import pandas as pd
import glob, os

# Load the tile encoder
tile_encoder = gigapath.load_tile_encoder("hf_hub:prov-gigapath/prov-gigapath")
tile_encoder.eval()

transform = gigapath.tile_transform(img_size=256)

def encode_tiles(svs_path, output_csv):
    tiles = gigapath.tile_slide(svs_path, mpp=0.5, tile_size=256)
    embeddings = []
    for tile_img, coords in tiles:
        t = transform(tile_img).unsqueeze(0)
        with torch.no_grad():
            emb = tile_encoder(t).squeeze().numpy()
        embeddings.append({"x": coords[0], "y": coords[1], **{f"d{i}": v for i, v in enumerate(emb)}})
    pd.DataFrame(embeddings).to_csv(output_csv, index=False)
```

Output: one CSV per slide, rows = tiles, columns = `x`, `y`, `d0`…`d1535`.

# Running Slide-Level Extraction

The slide encoder aggregates all tile embeddings using LongNet attention.

```python
slide_encoder = gigapath.load_slide_encoder("hf_hub:prov-gigapath/prov-gigapath")
slide_encoder.eval()

def encode_slide(tile_csv, output_csv):
    df = pd.read_csv(tile_csv)
    coords = torch.tensor(df[["x", "y"]].values, dtype=torch.float32)
    tile_embs = torch.tensor(df[[f"d{i}" for i in range(1536)]].values,
                             dtype=torch.float32).unsqueeze(0)
    with torch.no_grad():
        slide_emb = slide_encoder(tile_embs, coords.unsqueeze(0)).squeeze().numpy()
    pd.DataFrame([slide_emb]).to_csv(output_csv, index=False)
```

Output: one CSV per slide, a single row of 1,536 values.

# Batch Processing on a Compute Cluster

For large-scale processing across all TCGA slides:

```bash
#!/bin/bash
# Run on a SLURM cluster, one job per slide

apptainer run \
    --nv \
    --env HF_TOKEN="${HF_TOKEN}" \
    --bind /data/tcga:/data/tcga \
    --bind /data/output:/data/output \
    gigapath.sif \
    python extract_features.py \
        --input  /data/tcga/${FILE_ID}.svs \
        --output /data/output/${FILE_ID}_tiles.csv \
        --level  tile
```

# Accessing Pre-Computed Embeddings

If you do not have GPU resources, the pre-computed embeddings for all TCGA
slides are available via `imageFeatureTCGA`:

```{r access, eval=FALSE}
library(imageFeatureTCGA)
library(dplyr)

# Tile-level embeddings
getCatalog("provgigapath") |>
    dplyr::filter(level == "tile_level", Project.ID == "TCGA-OV") |>
    head(1) |>
    getFileURLs() |>
    ProvGiga() |>
    import()

# Slide-level embeddings
getCatalog("provgigapath") |>
    dplyr::filter(level == "slide_level", Project.ID == "TCGA-OV") |>
    head(1) |>
    getFileURLs() |>
    ProvGiga() |>
    import()
```

# Session Information

```{r session-info, eval=TRUE}
sessioninfo::session_info()
```
