---
title: "EuroBioC2026 Workshop: Histopathology Image Analysis in R/Bioconductor"
subtitle: "From TCGA images to multi-modal integration"
author:
  - name: Ilaria Billato
    affiliation: Department of Biology, University of Padova
    email: ilaria.billato@unipd.it
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: false
vignette: >
  %\VignetteIndexEntry{EuroBioC2026 Workshop}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

---

> **EuroBioC2026** | June 3–4, 2026 (TBD) | 90 minutes

---

# Workshop Description

Histopathological images provide unparalleled insights into tissue architecture,
cellular morphology, and tumor spatial organization. While these images are
routinely used in cancer research and clinical practice, their computational
analysis typically relies on specialized software outside the R/Bioconductor
ecosystem.

This workshop introduces the **imageTCGA ecosystem** — a suite of four
Bioconductor packages that bridges histopathological image analysis with
multi-omics integration. Using a single ovarian cancer (TCGA-OV) whole-slide
image as a running example, participants will explore the full analysis workflow:
from interactive data discovery to spatial statistics and multi-modal integration.

We applied a standardized, large-scale image analysis workflow to all TCGA
diagnostic WSIs (11,765 images, 9,640 cases, 32 cancer types) and release the
resulting pre-computed features as a ready-to-use resource within R/Bioconductor,
in compatible data structures: `SpatialExperiment`, `SpatialFeatureExperiment`,
and `MultiAssayExperiment`.

# Instructors

| Name | Affiliation |
|---|---|
| Ilaria Billato | Department of Biology, University of Padova |

# Pre-requisites

**Knowledge:**

- Basic R and Bioconductor familiarity
- Familiarity with `SummarizedExperiment` or `SpatialExperiment` is helpful
  but not required

**Software**:

```{r install, eval=FALSE}
docker run -e PASSWORD=bioc -p 8787:8787 ghcr.io/billila/imagetcgaworkflow:latest
```

This will start an RStudio Server instance accessible at http://localhost:8787 (username: rstudio, password: bioc)

# Workshop Participation

Participants will follow along with hands-on R code using pre-computed features
(no GPU or large downloads required). The workshop uses a single ovarian cancer
slide (`TCGA-23-1021`) as a running example throughout all sections.

Slides will be used to introduce concepts before each hands-on block.

# R/Bioconductor Packages Used

| Package | Role |
|---|---|
| `imageTCGA` | Interactive data discovery via Shiny app |
| `imageFeatureTCGA` | Import HoVerNet and Prov-GigaPath features |
| `imageTCGAutils` | Spatial statistics (PCA, Moran's I, LISA) |
| `HistoImagePlot` | Overlay segmentation on tissue thumbnails |
| `SpatialExperiment` | Bioconductor spatial data structure |
| `GenomicDataCommons` | Raw WSI download from GDC (demo only) |
| `dplyr` | Data filtering and manipulation |
| `ggplot2` | Visualization |
| `spdep` | Spatial weights and statistics |

# Time Outline

| Time | Topic | Format |
|---|---|---|
| 0–15 min | Introduction: histopathology + the Bioconductor gap | Slides |
| 15–30 min | Ecosystem overview: the 4 packages and the TCGA resource | Slides + Hands-on |
| 30-40 min | Raw Image Download using 'imageTCGA' |Hands-on |
| 40–55 min | HoVerNet nuclei features: import, explore, visualize | Hands-on |
| 55–70 min | Prov-GigaPath embeddings: tile & slide level, PCA, spatial stats | Hands-on |
| 70–80 min | Multi-modal integration with MOFA | Demo / Slides |
| 80–90 min | Q&A and free exploration | Open |

# Learning Goals

By the end of this workshop, participants will understand:

1. What pre-computed histopathology features are available for TCGA and how to
   access them without GPU or raw image downloads
2. How HoVerNet nuclei segmentation and Prov-GigaPath embeddings are structured
   as Bioconductor objects
3. How to perform spatial statistics on tile- and nucleus-level image features
4. How image-derived features can be integrated with TCGA molecular data

# Learning Objectives

Participants will be able to:

1. Use `getCatalog()` to browse the image feature catalogue and filter by cancer
   type, pipeline, and file format
2. Import HoVerNet nuclei data into a `SpatialExperiment` using
   `HoverNet() |> import()` and explore `colData()` and `spatialCoords()`
3. Visualize nuclei segmentation on a tissue thumbnail using
   `plotHoverNetH5ADOverlay()`
4. Import Prov-GigaPath tile- and slide-level embeddings using
   `ProvGiga() |> import()` and `ProvGigaList() |> import()`
5. Run PCA on tile embeddings and compute Moran's I with `spdep`

---

# Part 1 — Introduction (0–15 min) {.slides}

> **Format:** Slides

Key points to cover:

- Histopathological H&E images as the standard in clinical oncology
- The computational gap: image analysis tools are mostly outside R/Bioconductor
- TCGA as a unique resource: 11,765 WSIs across 32 cancer types
- Our approach: pre-compute once, release as Bioconductor-compatible objects

# Part 2 — Ecosystem Overview (15–30 min) {.slides}

> **Format:** Slides + Live demo (instructor runs the Shiny app)

```{r shiny}
library(imageTCGA)
imageTCGA()
```

Show participants how to:

1. Select **Ovarian Serous Cystadenocarcinoma (TCGA-OV)** in the Shiny app
2. Browse slides by patient, filter by metadata
3. Identify the example slide: `TCGA-23-1021-01Z-00-DX1`
4. Copy the GDC download code generated by the app

# Part 3 - Raw Image Download (30–40 min) {.slides}

> **Format:** Slides + code snippet (not run during workshop — files are ~1–2 GB)

```{r gdc-download}
# Example: download the raw WSI from GDC
# (this is what imageTCGA generates for you)
library(GenomicDataCommons)

# TBD: add actual GDC query for TCGA-23-1021
# file_ids <- c( "877ab00f-2431-4a99-86a7-53f39e233426")
# lapply(file_ids, gdcdata)
```

> Raw WSIs are H&E-stained SVS files (~1–2 GB each). For the rest of
> this workshop we use pre-computed features — no download required.

# Part 4 — HoVerNet Nuclei Features (40–55 min) {.handson}

> **Format:** Slides + Hands-on

```{r load}
library(imageFeatureTCGA)
library(HistoImagePlot)
library(SpatialExperiment)
library(dplyr)
library(ggplot2)
```

## Import

We use the H5AD file for the ovarian cancer example slide, which includes
additional per-nucleus features (mean intensity, nearest-neighbour distance):

HoVerNet data is available in JSON, GeoJSON, thumb and H5AD formats.

Note that the thumb format refers to the png thumbnails of the whole-slide images.
```{r explore imageFeatureTCGA}
getCatalog()

getCatalog("hovernet")
```

You can import HoVerNet segmentation results as either a SpatialExperiment or SpatialFeatureExperiment. Here we selectively import a file based on its filename, but you can also filter by other metadata fields such as Project.ID, pipeline, format, etc.

## HoVerNet outputs

HoVerNet provides nuclei segmentation and classification results at the
single-cell level. Each detected nucleus is represented by:

- spatial coordinates (`x`, `y`) in pixel units relative to the slide
- predicted cell type labels
- class probabilities
- polygon contours describing nuclear boundaries (when available)

When imported as a `SpatialExperiment` or `SpatialFeatureExperiment`,
the data are structured as follows:

- **columns** represent individual nuclei
- **colData** stores cell-level metadata (coordinates, cell types)
- **assays** may contain quantitative features (e.g., probabilities)
- **metadata** may include segmentation contours and image information

These objects enable spatial analyses and integration with other
Bioconductor workflows for spatial transcriptomics and imaging data.
```{r hovernet-import}
hspe <- getCatalog("hovernet") |>
    dplyr::filter(
        filename == paste(
            "TCGA-23-1021-01Z-00-DX1",
            "F07C221B-D401-47A5-9519-10DE59CA1E9D",
            "json",
            "gz",
            sep = "."
        )
    ) |>
    getFileURLs() |>
    HoverNet(outClass = "SpatialExperiment") |>
    import()
hspe
```

```{r}
colData(hspe)
```


## Explore

```{r hovernet-explore}
# How many nuclei? What cell types?
dim(hspe)
table(colData(hspe)$label)

# Spatial coordinates
head(spatialCoords(hspe))
```


```{r hovernet-explore h5ad}
# Per-nucleus features (H5AD only)
hspe_h5ad <- getCatalog("hovernet") |>
    dplyr::filter(
        filename == paste(
            "TCGA-23-1021-01Z-00-DX1",
            "F07C221B-D401-47A5-9519-10DE59CA1E9D",
            "h5ad",
            "gz",
            sep = "."
        )
    ) |>
    getFileURLs() |>
    HoverNet(outClass = "SpatialExperiment") |>
    import()
hspe_h5ad

head(colData(hspe_h5ad)[, c("mean_intensity", "nearest_neighbor_distance")])
```

## Visualize

```{r hovernet-plot}
thumb_path <- paste0(
    "https://store.cancerdatasci.org/hovernet/thumb/",
    "TCGA-23-1021-01Z-00-DX1.F07C221B-D401-47A5-9519-10DE59CA1E9D.png"
)

plotHoverNetH5ADOverlay(
    hspe_h5ad,
    thumb_path,
    title = "Ovarian Cancer — Cell Segmentation (TCGA-23-1021)"
)
```

# Part 5 — Prov-GigaPath Embeddings (55–70 min) {.handson}

> **Format:** Hands-on

```{r load-utils}
library(imageTCGAutils)
library(spdep)
```

## ProvGigaPath embeddings

ProvGigaPath is a foundation model trained on large-scale pathology image
tiles that produces high-dimensional embeddings summarizing visual and
morphological features.

Two levels of embeddings are provided:

### Slide-level embeddings

Slide-level embeddings summarize the entire whole-slide image into a
single feature vector.

- one row per slide
- embedding dimension corresponds to the encoder output size
- suitable for slide-level prediction or clustering tasks

### Tile-level embeddings

Tile-level embeddings provide localized representations of tissue regions.

Each tile entry includes:

- spatial coordinates (`tile_x`, `tile_y`) corresponding to the tile
  position on the slide
- a high-dimensional embedding vector
- optional metadata describing tile extraction parameters

These embeddings enable spatial analyses of tissue heterogeneity and can
be integrated with cell-level data from HoVerNet using complementary
packages such as `imageTCGAutils`.


## Import tile-level embeddings

## Slide-level embeddings

ProvGigaPath embeddings summarize tile or slide-level image features. In
this example, we import slide-level embeddings for a single file. Each
row corresponds to a slide, with an embedding vector describing the
image-derived features.

```{r explore imageFeatureTCGA provgiga}
getCatalog()

getCatalog("provgigapath")
```

```{r provgigapath slide level data}
pg_slide <- getCatalog("provgigapath") |>
    dplyr::filter(
        filename == paste(
            "TCGA-23-1021-01Z-00-DX1",
            "F07C221B-D401-47A5-9519-10DE59CA1E9D",
            "csv",
            "gz",
            sep = "."
        ) &
        level == "slide_level"
    ) |>
    getFileURLs() |>
    ProvGiga() |>
    import()
```

```{r provgigapath tile level data}
pg_tiles <- getCatalog("provgigapath") |>
    dplyr::filter(
        filename == paste(
            "TCGA-23-1021-01Z-00-DX1",
            "F07C221B-D401-47A5-9519-10DE59CA1E9D",
            "csv",
            "gz",
            sep = "."
        ) &
        level == "tile_level"
    ) |>
    getFileURLs() |>
    ProvGiga() |>
    import()

pg_tiles
dim(pg_tiles)
# 211 tiles × 1,543 columns (tile_x, tile_y + 1,536 embedding dims named 0–1535)
```



## PCA on tile embeddings

```{r prov-pca}
embedding_cols <- grep("^[0-9]+$", names(pg_tiles), value = TRUE)
coords <- pg_tiles[, c("tile_x", "tile_y")]

pca_res <- prcomp(pg_tiles[, embedding_cols], scale. = TRUE)
pca_df <- data.frame(coords, PC1 = pca_res$x[, 1], PC2 = pca_res$x[, 2])

ggplot(pca_df, aes(x = tile_x, y = tile_y, color = PC1)) +
    geom_point(size = 1) +
    scale_color_viridis_c() +
    coord_fixed() +
    theme_minimal() +
    labs(title = "Tile PCA — PC1 mapped to spatial position",
         subtitle = "TCGA-23-1021 (Ovarian Cancer)")
```

## Point Pattern Analysis

Point Pattern Analysis (PPA) provides a framework to investigate the spatial
distribution of cells within histopathology images. When HoVerNet output
is imported through `ImageFeatureTCGA`, each cell is represented with
spatial coordinates and associated morphological or intensity features.
These data allow the computation of global and local spatial autocorrelation
metrics, such as Moran's I, Local Moran's I (LISA), Geary's C,
and Local Spatial Heteroscedasticity (LOSH).

Here, we illustrate how to perform spatial autocorrelation
analysis starting from a `SpatialFeatureExperiment` object created from
HoVerNet segmentation results.

```{r}
hsfe <- getCatalog("hovernet") |>
    dplyr::filter(
        filename == paste(
            "TCGA-23-1021-01Z-00-DX1",
            "F07C221B-D401-47A5-9519-10DE59CA1E9D",
            "h5ad",
            "gz",
            sep = "."
        )
    ) |>
    getFileURLs() |>
    HoverNet(outClass = "SpatialFeatureExperiment") |>
    import()
hsfe
```

To investigate spatial patterns in the tissue, we use the PCA-reduced embeddings
for each tile. Each tile has a physical location (tile_x, tile_y) on the slide,
which allows us to explore how similar embedding values cluster across space. 
We construct a k-nearest neighbor graph to define which tiles are spatially 
“connected,” and then compute global and local spatial autocorrelation metrics. 

## Spatial autocorrelation (Moran's I)

Moran's I quantifies the degree to which similar values cluster spatially. 
A positive Moran's I indicates spatial clustering, while negative values 
indicate spatial dispersion.


```{r coords}
pg_tiles <- bind_cols(
    pg_tiles,
    as_tibble(pca_res$x)[, 1:2] |> rename(PC1 = "PC1", PC2 = "PC2")
)

coords <- pg_tiles[, c("tile_x", "tile_y")]
```



```{r morans-i}
nb <- knn2nb(knearneigh(coords, k = 6))
lw <- nb2listw(nb, style = "W")
```

Next, we calculate global spatial autocorrelation using Moran’s I and Geary’s C,
which quantify the overall tendency of similar PC1 values to cluster or disperse
on the tissue slide. We also compute Local Moran’s I (LISA) to detect local 
clusters of similar embedding values.

```{r metrics}
mi <- moran.test(pg_tiles$PC1, lw)
gc <- geary.test(pg_tiles$PC1, lw)
lisa <- localmoran(pg_tiles$PC1, lw)
pg_tiles$localI <- lisa[, "Ii"]
pg_tiles$localI_pval <- lisa[, "Pr(z != E(Ii))"]

mi
gc
```

We visualize the spatial patterns. The Moran scatterplot shows the relationship 
between each tile’s PC1 value and the mean of its neighbors, while the LISA plot
highlights local clusters (“hotspots”) of high or low PC1 values across the 
tissue slide.

```{r metrics 2}
moran.plot(pg_tiles$PC1, lw, labels = FALSE,
                main = "Moran scatterplot of PC1")

# LISA visualization
df_lisa <- data.frame(coords, Ii = lisa[, "Ii"])
ggplot(df_lisa, aes(x = tile_x, y = tile_y, color = Ii)) +
    geom_point(size = 1) +
    scale_color_viridis_c() +
    coord_equal() +
    theme_minimal() +
    ggtitle("Local Moran's I (LISA) for PC1")
```

- **Ripley's K / L function** — tests for clustering vs. regularity at
  multiple spatial scales
- **Cross-type K function** — spatial interaction between pairs of cell types
  (e.g. neoplastic and inflammatory)
  
```{r}
library(spatstat)

coords <- as.data.frame(spatialCoords(hspe))
types  <- colData(hspe)$label

# Create a marked point pattern
win <- owin(
  xrange = range(coords$x),
  yrange = range(coords$y)
)
pp <- ppp(x = coords$x, y = coords$y, marks = factor(types), window = win)

# Ripley's L for neoplastic cells
L_neo <- Lest(subset(pp, marks == "neopla"), correction = "isotropic")
plot(L_neo, main = "Ripley's L — neoplastic nuclei")

# Cross-type K: neoplastic vs. inflammatory - run if you have enough RAM (>20GB)
# K_cross <- Kcross(pp, i = "neopla", j = "inflam", correction = "isotropic")
# plot(K_cross, main = "Cross-type K: neoplastic vs. inflammatory")
```


## Adding HoverNet Nuclei Features

You can import HoVerNet segmentation results as a
`SpatialExperiment` or `SpatialFeatureExperiment`.

In this section we want show you how to integrate HoVerNet classification and 
segmentation output with Prov-GigaPath embeddings.

```{r importHover}
# import HoVerNet
hspe_h5ad

# import Prov-GigaPath
pg_tiles
```



```{r cell coordinates}
# Extract cell coordinates from HoVerNet
cell_coords <- spatialCoords(hspe_h5ad)

# Extract nuclei metadata 
cell_meta <- colData(hspe_h5ad)
cell_meta$x <-cell_coords[,1]
cell_meta$y <-cell_coords[,2]
```

Visualizing Hovernet nuclei vs tile coordinates to see that they do not match 
perfectly. 

You can use 'matchHoverNetToTiles' to compute the scaling factor.

```{r plotting hovernet vs tiles PCA}
plot(cell_meta$x, cell_meta$y, pch=16, col="#0000FF20")
points(pg_tiles$tile_x, 
        pg_tiles$tile_y, 
        pch=16, 
        col="#FF000020")

```


## Scale factor between nuclei coordinates and tile coordinates

```{r compute scaling factor}
match_hv_pg <- matchHoverNetToTiles(hspe_h5ad, pg_tiles)
```

```{r visualizing tile level with hovernet}
# ggplot(match_hv_pg$tiles_with_nuclei, aes(tile_x, tile_y, 
#                                     color = cell_type_label, 
#                                     size = N)) +
#     geom_point(alpha = 0.7) +
#     coord_equal() +
#     theme_minimal() +
#     labs(title = "All HoverNet cell types per tile")

ggplot(match_hv_pg$tiles_with_nuclei, aes(tile_x, tile_y, 
                                    color = dominant_cell_type)) +
    geom_point(size = 1) +
    coord_equal() +
    theme_minimal() +
    labs(title = "Per-tile dominant HoverNet cell type")
```


# Part 6 — Multi-Modal Integration (70–80 min) {.demo}

> **Format:** Slides / Hands-on

Planned content:

- Combine Prov-GigaPath slide-level embeddings with TCGA RNA-seq or
  clinical data for TCGA-OV
- Run MOFA+ (`MOFA2`) to identify joint factors of variation
- Visualize latent factors coloured by clinical annotation

```{r mofa-demo}
library(MOFA2)

```

# Load data

```{r load embeddings}
hspe

```


# Part 7 — Q&A and Free Exploration (80–90 min)

> **Format:** Open

Participants can explore their cancer type of interest!


---

# Session Information

```{r session-info, eval=TRUE}
sessioninfo::session_info()
```
