---
title: "EuroBioC2026 Workshop: Histopathology Image Analysis in R/Bioconductor"
subtitle: "From TCGA images to multi-modal integration"
author:
  - name: Ilaria Billato
    affiliation: Department of Biology, University of Padova
    email: ilaria.billato@unipd.it
date: "`r Sys.Date()`"
package: imageTCGAWorkflow
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: false
vignette: >
  %\VignetteIndexEntry{EuroBioC2026 Workshop}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

---

> **EuroBioC2026** | June 3–4, 2026 (TBD) | 90 minutes

---

# Workshop Description

Histopathological images provide unparalleled insights into tissue architecture,
cellular morphology, and tumor spatial organization. While these images are
routinely used in cancer research and clinical practice, their computational
analysis typically relies on specialized software outside the R/Bioconductor
ecosystem.

This workshop introduces the **imageTCGA ecosystem** — a suite of four
Bioconductor packages that bridges histopathological image analysis with
multi-omics integration. Using a single ovarian cancer (TCGA-OV) whole-slide
image as a running example, participants will explore the full analysis workflow:
from interactive data discovery to spatial statistics and multi-modal integration.

We applied a standardized, large-scale image analysis workflow to all TCGA
diagnostic WSIs (11,765 images, 9,640 cases, 32 cancer types) and release the
resulting pre-computed features as a ready-to-use resource within R/Bioconductor,
in compatible data structures: `SpatialExperiment`, `SpatialFeatureExperiment`,
and `MultiAssayExperiment`.

# Instructors

| Name | Affiliation |
|---|---|
| Ilaria Billato | Department of Biology, University of Padova |

# Pre-requisites

**Knowledge:**

- Basic R and Bioconductor familiarity
- Familiarity with `SummarizedExperiment` or `SpatialExperiment` is helpful
  but not required

**Software** (install before the workshop):

```{r install, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install(c(
    "billila/imageTCGA",
    "waldronlab/imageFeatureTCGA",
    "waldronlab/imageTCGAutils",
    "waldronlab/HistoImagePlot",
    "SpatialExperiment",
    "BiocFileCache",
    "dplyr",
    "ggplot2",
    "spdep"
))
```

# Workshop Participation

Participants will follow along with hands-on R code using pre-computed features
(no GPU or large downloads required). The workshop uses a single ovarian cancer
slide (`TCGA-23-1021`) as a running example throughout all sections.

Slides will be used to introduce concepts before each hands-on block.

# R/Bioconductor Packages Used

| Package | Role |
|---|---|
| `imageTCGA` | Interactive data discovery via Shiny app |
| `imageFeatureTCGA` | Import HoVerNet and Prov-GigaPath features |
| `imageTCGAutils` | Spatial statistics (PCA, Moran's I, LISA) |
| `HistoImagePlot` | Overlay segmentation on tissue thumbnails |
| `SpatialExperiment` | Bioconductor spatial data structure |
| `GenomicDataCommons` | Raw WSI download from GDC (demo only) |
| `dplyr` | Data filtering and manipulation |
| `ggplot2` | Visualization |
| `spdep` | Spatial weights and statistics |

# Time Outline

| Time | Topic | Format |
|---|---|---|
| 0–10 min | Introduction: histopathology + the Bioconductor gap | Slides |
| 10–20 min | Ecosystem overview: the 4 packages and the TCGA resource | Slides |
| 20–30 min | Data discovery with `imageTCGA` (Shiny app) | Live demo |
| 30–40 min | Raw image download with `GenomicDataCommons` | Slides + snippet |
| 40–55 min | HoVerNet nuclei features: import, explore, visualize | Hands-on |
| 55–70 min | Prov-GigaPath embeddings: tile & slide level, PCA, spatial stats | Hands-on |
| 70–80 min | Multi-modal integration with MOFA | Demo / Slides |
| 80–90 min | Q&A and free exploration | Open |

# Learning Goals

By the end of this workshop, participants will understand:

1. What pre-computed histopathology features are available for TCGA and how to
   access them without GPU or raw image downloads
2. How HoVerNet nuclei segmentation and Prov-GigaPath embeddings are structured
   as Bioconductor objects
3. How to perform spatial statistics on tile- and nucleus-level image features
4. How image-derived features can be integrated with TCGA molecular data

# Learning Objectives

Participants will be able to:

1. Use `getCatalog()` to browse the image feature catalogue and filter by cancer
   type, pipeline, and file format
2. Import HoVerNet nuclei data into a `SpatialExperiment` using
   `HoverNet() |> import()` and explore `colData()` and `spatialCoords()`
3. Visualize nuclei segmentation on a tissue thumbnail using
   `plotHoverNetH5ADOverlay()`
4. Import Prov-GigaPath tile- and slide-level embeddings using
   `ProvGiga() |> import()` and `ProvGigaList() |> import()`
5. Run PCA on tile embeddings and compute Moran's I with `spdep`

---

# Part 1 — Introduction (0–10 min) {.slides}

> **Format:** Slides

Key points to cover:

- Histopathological H&E images as the standard in clinical oncology
- The computational gap: image analysis tools are mostly outside R/Bioconductor
- TCGA as a unique resource: 11,765 WSIs across 32 cancer types
- Our approach: pre-compute once, release as Bioconductor-compatible objects

# Part 2 — Ecosystem Overview (10–20 min) {.slides}

> **Format:** Slides

The four packages and their roles:

| Package | Input | Output |
|---|---|---|
| `imageTCGA` | TCGA metadata | Shiny explorer + GDC download code |
| `imageFeatureTCGA` | Feature catalogue URLs | `SpatialExperiment`, `SummarizedExperiment` |
| `imageTCGAutils` | SPE / SE objects | PCA, Moran's I, Geary's C, LISA, `matchHoverNetToTiles()` |
| `HistoImagePlot` | SPE + thumbnail URL | Segmentation overlay plots |

The TCGA resource:

- **HoVerNet:** 33,177 files (JSON, GeoJSON, H5AD, PNG thumbnails)
- **Prov-GigaPath:** 21,076 files (tile-level and slide-level CSV)

# Part 3 — Data Discovery with `imageTCGA` (20–30 min) {.demo}

> **Format:** Live demo (instructor runs the Shiny app)

```{r shiny}
library(imageTCGA)
imageTCGA()
```

Show participants how to:

1. Select **Ovarian Serous Cystadenocarcinoma (TCGA-OV)** in the Shiny app
2. Browse slides by patient, filter by metadata
3. Identify the example slide: `TCGA-23-1021-01Z-00-DX1`
4. Copy the GDC download code generated by the app

# Part 4 — Raw Image Download (30–40 min) {.slides}

> **Format:** Slides + code snippet (not run during workshop — files are ~1–2 GB)

```{r gdc-download}
# Example: download the raw WSI from GDC
# (this is what imageTCGA generates for you)
library(GenomicDataCommons)

# TBD: add actual GDC query for TCGA-23-1021
# file_id <- "F07C221B-D401-47A5-9519-10DE59CA1E9D"
# gdcdata(file_id, destination_dir = "wsi/")
```

> Raw WSIs are H&E-stained TIFF/SVS files (~1–2 GB each). For the rest of
> this workshop we use pre-computed features — no download required.

# Part 5 — HoVerNet Nuclei Features (40–55 min) {.handson}

> **Format:** Hands-on

```{r load}
library(imageFeatureTCGA)
library(HistoImagePlot)
library(SpatialExperiment)
library(dplyr)
library(ggplot2)
```

## Import

We use the H5AD file for the ovarian cancer example slide, which includes
additional per-nucleus features (mean intensity, nearest-neighbour distance):

```{r hovernet-import}
hov_file <- paste0(
    "https://store.cancerdatasci.org/hovernet/h5ad/",
    "TCGA-23-1021-01Z-00-DX1.F07C221B-D401-47A5-9519-10DE59CA1E9D.h5ad.gz"
)

hn_spe <- HoverNet(hov_file, outClass = "SpatialExperiment") |>
    import()

hn_spe
```

Alternatively, import via the catalogue:

```{r hovernet-catalog}
hn_spe <- getCatalog("hovernet") |>
    dplyr::filter(
        tcga_barcode == "TCGA-23-1021-01Z-00-DX1.F07C221B-D401-47A5-9519-10DE59CA1E9D",
        format == "h5ad"
    ) |>
    getFileURLs() |>
    HoverNet(outClass = "SpatialExperiment") |>
    import()
```

## Explore

```{r hovernet-explore}
# How many nuclei? What cell types?
dim(hn_spe)
table(colData(hn_spe)$label)

# Spatial coordinates
head(spatialCoords(hn_spe))

# Per-nucleus features (H5AD only)
head(colData(hn_spe)[, c("mean_intensity", "nearest_neighbor_distance")])
```

## Visualize

```{r hovernet-plot}
thumb_path <- paste0(
    "https://store.cancerdatasci.org/hovernet/thumb/",
    "TCGA-23-1021-01Z-00-DX1.F07C221B-D401-47A5-9519-10DE59CA1E9D.png"
)

plotHoverNetH5ADOverlay(
    hn_spe,
    thumb_path,
    title = "Ovarian Cancer — Cell Segmentation (TCGA-23-1021)"
)
```

# Part 6 — Prov-GigaPath Embeddings (55–70 min) {.handson}

> **Format:** Hands-on

```{r load-utils}
library(imageTCGAutils)
library(spdep)
```

## Import tile-level embeddings

```{r prov-tile}
tile_url <- paste0(
    "https://store.cancerdatasci.org/provgigapath/tile_level/",
    "TCGA-23-1021-01Z-00-DX1.F07C221B-D401-47A5-9519-10DE59CA1E9D.csv.gz"
)

pg_tiles <- ProvGiga(tile_url) |> import()

# 211 tiles × 1,543 columns (tile_x, tile_y + 1,536 embedding dims named 0–1535)
dim(pg_tiles)
```

## PCA on tile embeddings

```{r prov-pca}
embedding_cols <- grep("^[0-9]+$", names(pg_tiles), value = TRUE)
coords <- pg_tiles[, c("tile_x", "tile_y")]

pca_res <- prcomp(pg_tiles[, embedding_cols], scale. = TRUE)
pca_df <- data.frame(coords, PC1 = pca_res$x[, 1], PC2 = pca_res$x[, 2])

ggplot(pca_df, aes(x = tile_x, y = tile_y, color = PC1)) +
    geom_point(size = 3) +
    scale_color_viridis_c() +
    coord_fixed() +
    theme_minimal() +
    labs(title = "Tile PCA — PC1 mapped to spatial position",
         subtitle = "TCGA-23-1021 (Ovarian Cancer)")
```

## Spatial autocorrelation (Moran's I)

```{r morans-i}
nb <- knn2nb(knearneigh(as.matrix(coords), k = 6))
lw <- nb2listw(nb, style = "W")

mi <- moran.test(pca_df$PC1, lw)
mi
```

## Import slide-level embeddings for TCGA-OV

```{r prov-slide}
# Multiple slides → SummarizedExperiment (768 × n slides)
pgl <- getCatalog("provgigapath") |>
    dplyr::filter(level == "slide_level", Project.ID == "TCGA-OV") |>
    dplyr::slice(1:10) |>
    getFileURLs() |>
    ProvGigaList() |>
    import()

pgl
# class: SummarizedExperiment
# dim: 768 10
# assays(1): embeddings
```

# Part 7 — Multi-Modal Integration (70–80 min) {.demo}

> **Format:** Demo / Slides — TBD (working code to be added before the workshop)

Planned content:

- Combine Prov-GigaPath slide-level embeddings with TCGA RNA-seq or
  clinical data for TCGA-OV
- Run MOFA+ (`MOFA2`) to identify joint factors of variation
- Visualize latent factors coloured by clinical annotation

```{r mofa-demo}
# TBD: MOFA2 integration example with TCGA-OV
# library(MOFA2)
# ...
```

> See the [Downstream Analyses vignette](downstream-analyses.html) for a
> more detailed walkthrough.

# Part 8 — Q&A and Free Exploration (80–90 min)

> **Format:** Open

Participants can explore their cancer type of interest:

```{r free-explore}
# Browse available cancer types
getCatalog("hovernet") |>
    dplyr::count(Project.ID, sort = TRUE)

# Pick a cancer type and repeat the workflow
getCatalog("provgigapath") |>
    dplyr::filter(level == "slide_level", Project.ID == "TCGA-BRCA") |>
    dplyr::slice(1:5) |>
    getFileURLs() |>
    ProvGigaList() |>
    import()
```

---

# Session Information

```{r session-info, eval=TRUE}
sessioninfo::session_info()
```
